---
title: "Evaluation of model in gse90496"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Introduction

In this model, we will evaluate the first configuration of the CNN model to classify TCGA samples in the GEO dataset GSE90496. 

```{r}
library(tidyverse)
library(HDF5Array)
library(SummarizedExperiment)

version <- "2021-05-27"

classification <- read_table2(paste0("../results/gse90496/", version, "/prediction.tsv"))
SE <- loadHDF5SummarizedExperiment(dir = paste0("../results/preprocess_gse90496/", version), prefix = "GSE90496_rawinputProbes_")
```

## TCGA trained model

```{r}
tcga_class <- classification$`0`
class_df <- data.frame(ori_class = SE$`methylation class:ch1`, tcga_class) %>%
  tibble() %>%
  mutate(ori_class_sum = ifelse(grepl("^CONTR", ori_class), "Control", 
                                ifelse(grepl("GBM", ori_class), "GBM", "Tumor")),
         match = (ori_class_sum == "Control" & tcga_class == "Normal") |
           (ori_class_sum == "GBM" & tcga_class == "TCGA-GBM") |
           (ori_class_sum == "Tumor" & tcga_class == "TCGA-LGG")
  )
map_df <- table(class_df$ori_class, class_df$tcga_class) %>%
  data.frame() %>%
  tibble() %>%
  mutate(GEO = Var1,
         model = Var2) 
```

```{r fig.height = 5, fig.width = 20, fig.align = "center"}

map_df %>%
    ggplot(aes(x = GEO, y = model, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", Freq)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

```{r fig.height = 5, fig.width = 20, fig.align = "center"}
map_df %>%
  filter(!Var2 %in% c("TCGA-ACC", "TCGA-BLCA", "TCGA-LUAD", "TCGA-PAAD", "TCGA-PRAD", "TCGA-THCA")) %>%
    ggplot(aes(x = GEO, y = model, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", Freq)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

Most of the samples are classified in the TCGA-LGG category (Brain Lower Grade Glioma), while other group of samples are classified under the TCGA-GBM category (Glioblastoma multiforme). Some control samples were also correctly identified as controls. However, some tumor subtypes were only classified as normal.

```{r}
table(class_df$tcga_class, class_df$ori_class_sum)
```

```{r}
class_df %>%
  summarize(accuracy = mean(match),
            recall = sum(match & ori_class_sum != "Control") / sum(ori_class_sum != "Control"),
            precision = sum(tcga_class %in% c("TCGA-GBM", "TCGA-LGG") &  ori_class_sum != "Control") /  sum(tcga_class %in% c("TCGA-GBM", "TCGA-LGG")),
            specificity = sum(match & ori_class_sum == "Control") / sum(ori_class_sum == "Control"))

```
We achieved an accuracy of 72.5% to classify tumor samples in the new cohort. Our model is lacking specificity, as it is not able to identify control samples. 


## Transfer learning

```{r}
confusion <- read.delim(paste0("../results/gse90496_transfer/", version, "/confussionMatrix.tsv"))
report <- read_table(paste0("../results/gse90496_transfer/", version, "/classificationReport.txt"))
eval <- read_table2(paste0("../results/gse90496_transfer/", version, "/training_evaluation.tsv"), col_names =  FALSE)
class <- read.delim(paste0("../results/gse90496_transfer/", version, "/prediction.tsv"))
```

### Training evaluation

```{r}
eval_df <- eval[, -1] %>%
  t() %>%
  data.frame() %>%
  tibble()
colnames(eval_df) <- eval$X1


eval_df2 <- eval_df %>%
  mutate(Epoch = seq_len(nrow(eval_df))) %>%
  gather(Category, Score, 1:4)

eval_df2 %>%
  filter(Category %in% c("val_loss", "loss")) %>%
  mutate(Dataset = ifelse(Category == "loss", "Training", "Validation")) %>%
  ggplot(aes(x = Epoch, y = Score, group = Dataset, color = Dataset)) +
  geom_line() +
  ylab("Loss") +
  theme_bw()
```

Loss function drops consistently for all epochs.

```{r}
eval_df2 %>%
  filter(Category %in% c("categorical_accuracy", "val_categorical_accuracy")) %>%
  mutate(Dataset = ifelse(Category == "categorical_accuracy", "Training", "Validation")) %>%
  ggplot(aes(x = Epoch, y = Score, group = Dataset, color = Dataset)) +
  geom_line() +
  ylab("Accuracy") +
  theme_bw()
```

Accuracy follows the same pattern than loss, with a constant increasing along the whole range.

### Classification

```{r}
tail(report, 3)
```

We obtained an accuracy and recall around 80%.

```{r}
filter(report, !X1 %in% c("accuracy", "macro avg", "weighted avg")) %>%
  arrange(precision)
```


```{r fig.height = 11, fig.width = 11, fig.align = "center"}
confusion %>%
  mutate(True = colnames(.)) %>%
  gather(Predicted, N, 1:91) %>%
  ggplot(aes(x = Predicted, y = True, fill = N)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", N)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```
Most of the are correctly classified. In order to explore the possible problems, we will focus on those tumors having more than 5 individuals wrongly classified:

```{r}
conf <- data.matrix(confusion)
diag(conf) <- 0
col <- which(colSums(conf) > 5)
row <- which(rowSums(conf) > 5)
index <- sort(unique(c(col, row)))

confusion[index, index] %>%
  mutate(True = colnames(.)) %>%
  gather(Predicted, N, seq_len(length(index))) %>%
  ggplot(aes(x = Predicted, y = True, fill = N)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", N)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

```{r}
class_mod <- class %>% 
  mutate(Real_fam = sapply(strsplit(Real, ","), `[`, 1),
         Predicted_fam = sapply(strsplit(Predicted, ","), `[`, 1))
class_mod %>%
  summarize(accuracy_family = mean(Real_fam == Predicted_fam),
            accuracy_tumor = mean(Real == Predicted))
```
