---
title: "Evaluation of GSE55763 model"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Introduction

In this model, we will evaluate the first configuration of the CNN model to classify blood samples by differences in sex.

```{r}
library(tidyverse)
version <- "2021-06-09"

confusion <- read_table2(paste0("../results/GSE55763_model/", version, "/confussionMatrix.tsv"))
report <- read_table(paste0("../results/GSE55763_model/", version, "/classificationReport.txt"))
eval <- read_table2(paste0("../results/GSE55763_model/", version, "/training_evaluation.tsv"), col_names =  FALSE)
```

## Training evaluation

```{r}
eval_df <- eval[, -1] %>%
  t() %>%
  data.frame() %>%
  tibble()
colnames(eval_df) <- eval$X1


eval_df2 <- eval_df %>%
  mutate(Epoch = seq_len(nrow(eval_df))) %>%
  gather(Category, Score, 1:4)

eval_df2 %>%
  filter(Category %in% c("val_loss", "loss")) %>%
  mutate(Dataset = ifelse(Category == "loss", "Training", "Validation")) %>%
  ggplot(aes(x = Epoch, y = Score, group = Dataset, color = Dataset)) +
  geom_line() +
  ylab("Loss") +
  theme_bw()
```

Loss function drops greatly between epochs 1 to 4 in both training and validation dataset. From epoch 5, it is slowly reduced, with a higher lost in training than in test, suggesting a possible overfitting.

```{r}
eval_df2 %>%
  filter(Category %in% c("categorical_accuracy", "val_categorical_accuracy")) %>%
  mutate(Dataset = ifelse(Category == "categorical_accuracy", "Training", "Validation")) %>%
  ggplot(aes(x = Epoch, y = Score, group = Dataset, color = Dataset)) +
  geom_line() +
  ylab("Accuracy") +
  theme_bw()
```

Accuracy follows the same pattern than loss, with a great improvement between epochs 1 to 4, and small improvements from this point. Higher increase in training accuracy compared with validation suggest a potential overfitting.

## Classification

```{r}
tail(report, 3)
```

A perfect accuracy.

```{r}
filter(report, !X1 %in% c("accuracy", "macro avg", "weighted avg")) %>%
  arrange(precision)
```

The sex of all samples was perfectly recalled.


```{r fig.height = 11, fig.width = 11, fig.align = "center"}
confusion %>%
  mutate(True = colnames(.)) %>%
  gather(Predicted, N, 1:2) %>%
  ggplot(aes(x = Predicted, y = True, fill = N)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", N)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

All samples are correctly predicted.