---
title: "Evaluation of model in gse90496"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Introduction

In this model, we will evaluate the first configuration of the CNN model to classify TCGA samples in the GEO dataset GSE90496. 

```{r}
library(tidyverse)
library(HDF5Array)
library(SummarizedExperiment)

version <- "2021-05-17"

classification <- read_table2(paste0("../results/gse90496/", version, "/prediction.tsv"))
confusion <- read_table2("../results/tcga_model/2021-05-14/confussionMatrix.tsv")
SE <- loadHDF5SummarizedExperiment(dir = paste0("../results/preprocess_gse90496/", version), prefix = "GSE90496_rawinputProbes_")
```

## TCGA trained model

```{r}
tcga_class <- colnames(confusion)[classification$`0` + 1]
class_df <- data.frame(ori_class = SE$`methylation class:ch1`, tcga_class) %>%
  tibble() %>%
  mutate(ori_class_sum = ifelse(grepl("CONTR", ori_class), "Control", 
                                ifelse(grepl("GBM", ori_class), "GBM", "Tumor")),
         match = (ori_class_sum == "Control" & tcga_class == "Normal") |
           (ori_class_sum == "GBM" & tcga_class == "TCGA-GBM") |
           (ori_class_sum == "Tumor" & tcga_class == "TCGA-LGG")
  )
map_df <- table(class_df$ori_class, class_df$tcga_class) %>%
  data.frame() %>%
  tibble() %>%
  mutate(GEO = Var1,
         model = Var2) 
```

```{r fig.height = 5, fig.width = 20, fig.align = "center"}

map_df %>%
    ggplot(aes(x = GEO, y = model, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", Freq)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

```{r fig.height = 5, fig.width = 20, fig.align = "center"}
map_df %>%
  filter(!Var2 %in% c("TCGA-ACC", "TCGA-BRCA", "TCGA-KIRP", "TCGA-MESO", "TCGA-UCEC")) %>%
    ggplot(aes(x = GEO, y = model, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%1.0f", Freq)), vjust = 1) +
   scale_fill_gradient(low = "white",
                      high = "blue") +
  theme(legend.position = "none",
        axis.text.x  = element_text(angle = 45, vjust=0.5))

```

Most of the samples are classified in the TCGA-LGG category (Brain Lower Grade Glioma), while other group of samples are classified under the TCGA-GBM category (Glioblastoma multiforme). None of the control samples was identified as control, although there are control samples in the dataset.

```{r}
table(class_df$tcga_class, class_df$ori_class_sum)
```

```{r}
class_df %>%
  summarize(accuracy = mean(match),
            recall = sum(match & ori_class_sum != "Control") / sum(ori_class_sum != "Control"),
            precision = sum(tcga_class %in% c("TCGA-GBM", "TCGA-LGG") &  ori_class_sum != "Control") /  sum(tcga_class %in% c("TCGA-GBM", "TCGA-LGG")),
            specificity = sum(match & ori_class_sum == "Control") / sum(ori_class_sum == "Control"))

```
We achieved an accuracy of 72.5% to classify tumor samples in the new cohort. Our model is lacking specificity, as it is not able to identify control samples. 
